---
title: "R Notebook"
output: html_notebook
---

```{r}
# Install packages
install.packages("stats")
install.packages("dendextend")

# Load in packages
library(tidyverse)
library(ggplot2)
library(lubridate)
library(skimr)
library(psych)
library(Hmisc)
library(corrplot)
library(naniar)
library(GGally)
library(janitor)
library(modelsummary)
library(jtools)
library(stats)
library(formattable)
library(dendextend)

# Load in data
df <- read_csv("cereal.csv")
```


PRELIM CHECK
```{r}
df %>% head()

df %>%
  gg_miss_var()
```


DESCRIPTIVES
```{r}
# Summary statistics - nutrient variables
df %>%
  select(calories, protein, fat, sodium, fiber, carbo, sugars, potass) %>%
  psych::describe()

# Summary statistics - other metric variables
df %>%
  select(weight, cups, rating) %>%
  psych::describe()
```

RENAME MANUFACTURER VALUES
```{r}
df <- df %>%
    mutate(mfr = recode(mfr,
                        "A" = "American Home Food Products",
                        "G" = "General Mills",
                        "K" = "Kelloggs",
                        "N" = "Nabisco",
                        "P" = "Post",
                        "Q" = "Quaker Oats",
                        "R" = "Ralston Purina"))
```

SUMMARY STATISTICS
```{r}

```

DESCRIPTIVES GROUPED BY VARIABLES
```{r}
# Group by manufacturer
## Mean - protein, fat, fiber, carbo, sugars
## Median - calories, sodium, potass
df %>%
  group_by(mfr) %>%
  summarise(
    mean_protein = round(mean(protein), digits = 3),
    mean_fat = round(mean(fat), digits = 3),
    mean_fiber = round(mean(fiber), digits = 3),
    mean_carbo = round(mean(carbo), digits = 3),
    mean_sugars = round(mean(sugars), digits = 3),
    med_calories = median(calories),
    med_sodium = median(sodium),
    med_potass = median(potass)) -> sum_table

# Reformat table
colnames(sum_table) <- c("Manufacturer", 
                         "Protein (Mean)",
                         "Fat (Mean)",
                         "Fiber (Mean)",
                         "Carbs (Mean)",
                         "Sugars (Mean)",
                         "Calories (Median)",
                         "Sodium (Median)",
                         "Potassium (Median)")

formattable(sum_table)

```

DISTRIBUTIONS & FREQUENCIES
```{r}
# Ratings
df %>%
  ggplot() + geom_histogram(aes(rating), bins = 25) +
  theme_bw() +
  labs(
    x = "Rating",
    y = "Number of Cereals"
  )


# Manufacturer
df %>%
  ggplot() + geom_bar(aes(mfr)) +
  theme_bw(base_size = 18) +
  labs(
    x = "Manufacturer",
    y = "Number of Cereal Brands"
  ) + coord_flip()


# Type (hot or cold) -- over 96% served cold
df %>% 
  tabyl(type) %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_totals()


# Vitamins -- over 80% at 25%
df %>% 
  tabyl(vitamins) %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_totals()


# Shelf -- 46.75% at shelf 3 (top); rest nearly evenly split at shelves 1 and 2
df %>% 
  tabyl(shelf) %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_totals()


# Ratings by manufacturer
df %>%
  group_by(mfr) %>%
  summarise(mean_rating = mean(rating)) %>%
  ggplot(aes(x = mfr, y = mean_rating)) + geom_bar(stat = "identity") +
  coord_flip() + theme_bw(base_size = 18) +
  labs(
    x = "Manufacturer",
    y = "Average Rating"
  )
  
```


DIMENSION REDUCTION


SCALE METRIC VARIABLES
```{r}
# protein, fat, sodium, fiber, carbo, sugars, potass
df %>%
  mutate(
    scale_protein = scale(protein),
    scale_fat = scale(fat),
    scale_sodium = scale(sodium),
    scale_fiber = scale(fiber),
    scale_carbo = scale(carbo),
    scale_sugars = scale(sugars),
    scale_potass = scale(potass),
    scale_calories = scale(calories)) -> df

```

VISUAL INSPECTION - CORRELATION MATRIX
```{r}
# Test 1
df %>%
  select(
    scale_protein, scale_fat, scale_sodium,
    scale_fiber, scale_carbo, scale_sugars, scale_potass,
    scale_calories) %>%
  as.matrix() %>% rcorr() -> df.corr

corrplot(df.corr$r, method = "shade", order = "alphabet", type = "lower", addCoef.col = TRUE, tl.srt = 45, tl.pos = "ld")


# Test 2 - take out sugars and fat and calories
df %>%
  select(
    scale_protein, scale_sodium,
    scale_fiber, scale_carbo, scale_potass) %>%
  as.matrix() %>% rcorr() -> df.corr_2

corrplot(df.corr_2$r, method = "shade", order = "alphabet", type = "lower", addCoef.col = TRUE, tl.srt = 45, tl.pos = "ld")

```


KMO, BARTLETT'S TEST
```{r}
# KMO
KMO(df.corr$r) # Correlation matrix 1
KMO(df.corr_2$r) # Correlation matrix 2 -- better score

# Bartlett's
cortest.bartlett(df.corr_2$r, n = 77) # Significant
```

DERIVING COMPONENTS
```{r}
# Scree plot with eigenvalue
PCA <- principal(df.corr_2$r)
plot(PCA$values, type = "lines")
lines(rep(1, 5), col = "red")

# Communalities
PCA <- principal(df.corr_2$r, nfactors = 2)
PCA$communality

```

PCA ROTATION
```{r}
# Orthogonal rotation - Varimax
PCArotate <- principal(df.corr_2$r, nfactors = 2, rotate = "varimax")

# Compare loadings to non-rotated solution
PCAnotrotate <- principal(df.corr_2$r, nfactors = 2, rotate = "none")

# Show loadings
PCAnotrotate$loadings
PCArotate$loadings

```

CREATE NEW VARIABLES
```{r}
# Use weights
w <- as.matrix(PCArotate$weights)

# Create labels and apply weights
df %>%
  mutate(
    BloodSugarHealth = cbind(df$scale_protein, df$scale_fiber, df$scale_potass, df$scale_sodium, df$scale_carbo) %*% w[,1],
    CarefulConsumption = cbind(df$scale_protein, df$scale_fiber, df$scale_potass, df$scale_sodium, df$scale_carbo) %*% w[,2]
  ) %>%
  {.} -> data.withpca

```

PLOT NEW VARIABLES (COMPONENTS)
```{r}
# Plot 2 new components
data.withpca %>%
  ggplot() + geom_point(aes(x = CarefulConsumption, y = BloodSugarHealth, color = factor(shelf))) +
  theme_bw(base_size = 18)

```


CLUSTERING


SELECT STANDARDIZED BASE VARIABLES
```{r}
# Scale/standardize 2 new variables for cluster input
data.withpca %>%
  mutate(
    scaledBloodSugarHealth = scale(BloodSugarHealth),
    scaledCarefulConsumption = scale(CarefulConsumption)
  ) %>%
  select(scaledBloodSugarHealth, scaledCarefulConsumption) %>% 
  {.} -> data.scaledPCA

```

CREATE A DISTANCE MATRIX
```{r}
dist_mat <- dist(data.scaledPCA, method = "euclidean")
```

CLUSTER USING WARD'S METHOD
```{r}
hclust_ward <- hclust(dist_mat, method = "ward.D")
```

PLOT DENDROGRAM
```{r}
# See number of segments -- choose 4 later
plot(hclust_ward) # 4 clusters

# Plot dendrogram with colors - 3 clusters
dend_obj <- as.dendrogram(hclust_ward)
color_dend <- color_branches(dend_obj, k = 3)
plot(color_dend)

# Plot dendrogram with colors - 4 clusters
dend_obj <- as.dendrogram(hclust_ward)
color_dend <- color_branches(dend_obj, k = 4)
plot(color_dend)

```

K-MEANS CLUSTERING
```{r}
# With starting point of 100, 3 clusters
data.scaledPCA %>%
  select(scaledBloodSugarHealth, scaledCarefulConsumption) %>% 
  kmeans(., 3, nstart = 100) -> kclust

# With starting point of 100, 4 clusters
data.scaledPCA %>%
  select(scaledBloodSugarHealth, scaledCarefulConsumption) %>% 
  kmeans(., 4, nstart = 100) -> kclust

# Add cluster as variable to data frame
data.withpca$Cluster <-kclust$cluster

```

PLOT CLUSTERS
```{r}
data.withpca %>%
  ggplot() + geom_point(aes(x = scaledCarefulConsumption, y = scaledBloodSugarHealth, color = factor(Cluster))) +
  theme_bw(base_size = 18) +
  labs(
    x = "Careful Consumption",
    y = "Blood Sugar Health"
  )

# Number of brands within each cluster
table(data.withpca$Cluster)

```

CLUSTER ANALYSIS - DESCRIPTOR VARIABLES
```{r}
# Average ratings grouped by cluster
data.withpca %>%
  group_by(Cluster) %>%
  summarise(mean_rating = mean(rating)) %>%
  ggplot(aes(x = Cluster, y = mean_rating)) + geom_bar(stat = "identity") +
  theme_bw(base_size = 18) +
  labs(
    x = "Cluster",
    y = "Average Cereal Rating"
  )

# Average calories grouped by cluster
data.withpca %>%
  group_by(Cluster) %>%
  summarise(mean_calories = mean(calories)) %>%
  ggplot(aes(x = Cluster, y = mean_calories)) + geom_bar(stat = "identity") +
  theme_bw(base_size = 18) +
  labs(
    x = "Cluster",
    y = "Average Cereal Calories"
  )

# Manufacturer frequency (preference) by cluster
data.withpca %>%
  select(
    Cluster,
    mfr) %>%
  group_by(Cluster) %>%
  ggplot() + geom_bar(aes(mfr)) + coord_flip() + facet_wrap(Cluster~., scales = "free") +
  theme_bw(base_size = 18) +
  labs(
    x = "Manufacturer",
    y = "Number of Brands"
  )

# Shelf frequency (preference) by cluster
data.withpca %>%
  select(
    Cluster,
    shelf) %>%
  group_by(Cluster) %>%
  ggplot() + geom_bar(aes(shelf)) + coord_flip() + facet_wrap(Cluster~., scales = "free") +
  theme_bw(base_size = 18) +
  labs(
    x = "Shelf",
    y = "Number of Brands"
  )

```

